---
title: "Text cleaning"
date: "2023-09-06"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

# Background

Sammy provided us with a couple of excel and text files documenting the types of drugs administered per-patient. We want to summarize the data. 

# Objectives

1. Summarize all the listed drugs
2. Heatmap of pre-categorized drug


# Pre-processing 

## Loading packages

```{r, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(readxl)
library(ggplot2)
library(reshape2)
library(stringr)

library(ComplexHeatmap)
```


## Directories

```{r}
wd <- list()
wd$main <- here()
wd$data <- file.path(wd$main, "data")
wd$dataP <- file.path(wd$data, "processed")
wd$output <- file.path(wd$main, "output")
wd$script <- file.path(wd$main, "scripts")
```


Create directories

```{r}
if (!file.exists(wd$output )) {
  dir.create(wd$output )
} else {
  print("Directory already exists")
}
```

Load functionm

```{r}
source(file.path(wd$script, "functions.R"))
```


# Objective 1

**Summarize all the listed drugs**

Read in the data

```{r}
dat <- read.delim(file.path(wd$data, "from_Sammy/meds-baseline.tsv"))
head(dat)
```

Each row represents a patient, with the 'record_id' column represeting the patient ID. We extract the 'current_meds_other' column. This is the column detailing what other medication was given. Since multiple drug is given in a column, we separate this.

Sometimes the drug is seperated by a ',' comma or a semi-colon ';'


```{r}
list_split <- str_split(dat$current_meds_other, ",|;")
names(list_split) <- dat$record_id


un_split <- unlist(list_split, use.names=TRUE)

df_split <- stack(list_split) %>%
  # Rename
  dplyr::rename(current_med = values,
                record_id = ind) 

# Remove any empty spaces at the start
df_split <- df_split %>%
  mutate(current_med = gsub("^ ", "", current_med),
         current_med = as.character(mapply(simpleCap, current_med))) %>%
  # Remove blanks
  filter(!current_med == "") %>%
  # Further modification
  mutate(
    # If the start is "-", remove the symbol
    current_med = gsub("^-", "", current_med),
    # Remove leading and trailing
    current_med = trimws(current_med, which="both", whitespace = "[ \t\r\n]"),
    current_med = trim(current_med)
    )

# # Remove some special symbol
# df_split2 <- df_split %>%
#   mutate(current_med = gsub(" ", " ", current_med))
# #write.csv(df_split2, "~/Desktop/tmp.csv")

```



Quick look at the harmonized category, we see some variation to the names, something like this. 

```{r}
filter(df_split, str_detect(current_med, "Hydrochlorothiazide")) %>% select(current_med) %>% unique()
filter(df_split, str_detect(current_med, "hydrochlorothiazide")) %>% select(current_med) %>% unique()
```

Lets further harmonize the name

```{r}
df_split2 <- df_split %>%
  mutate(
    current_med = 
      case_when(
        str_detect(current_med, "(?i)hydrochlorothiazide") ~ "Hydrochlorothiazide",
        str_detect(current_med, "(?i)Clonazepam") ~ "Clonazepam",
        str_detect(current_med, "(?i)cymbalta") ~ "Cymbalta",
        str_detect(current_med, "(?i)diphenhydramine") ~ "Diphenhydramine",
        str_detect(current_med, "(?i)gabapentin") ~ "Gabapentin",
        str_detect(current_med, "(?i)hydrocodone") ~ "Hydrocodone",
        str_detect(current_med, "(?i)imodium") ~ "Imodium",
        str_detect(current_med, "(?i)levofloxacin") ~ "Levofloxacin",
        str_detect(current_med, "(?i)ondansetron") ~ "Ondansetron",
        str_detect(current_med, "(?i)prochlorperazine") ~ "Prochlorperazine",
        str_detect(current_med, "(?i)soma") ~ "Soma",
        str_detect(current_med, "(?i)acetaminophen") ~ "Acetaminophen",    # SAME as oxydone?
        str_detect(current_med, "(?i)docusatesodium") ~ "Docusatesodium",
        str_detect(current_med, "(?i)methimazole") ~ "Methimazole",
        str_detect(current_med, "(?i)nicotine") ~ "Nicotine",
        str_detect(current_med, "(?i)oxycodone") ~ "Oxycodone",      # SAME as acetaminophen?
        str_detect(current_med, "(?i)prochlorperazine") ~ "Prochlorperazine",
        str_detect(current_med, "(?i)Coq10") ~ "Coq10",
        str_detect(current_med, "(?i)Aldactazide") ~ "Aldactazide",
        str_detect(current_med, "(?i)Buspirone") ~ "Buspirone",
        str_detect(current_med, "(?i)Cefdinir") ~ "Cefdinir",
        str_detect(current_med, "(?i)Potassium") ~ "Potassium",
        str_detect(current_med, "(?i)levothyroxine") ~ "Levothyroxine",
        str_detect(current_med, "(?i)nystatin") ~ "Nystatin",
        str_detect(current_med, "(?i)clearlax") ~ "Clearlax",
        str_detect(current_med, "(?i)Polyethylene") ~ "Polyethylene Glycol",
        #str_detect(current_med, "(?i)Albuterol") ~ "Albuterol",
        #str_detect(current_med, "(?i)Allopurinol") ~ "Allopurinol",
        
        
        

        
        
        
        
        
        
        TRUE ~ current_med),
    current_med = gsub("[0-9.-]+mg.*", "", current_med),
    # Remove "Mg" in capital letters with or without a space before a decimal number
    current_med = gsub("([A-Z] )?\\d+(\\.\\d+)? Mg.*", "", current_med)
      )

#str_detect(current_med, "(?i)") ~ "",
```

Things to check

Drug combination, what to do
  - Oxycodone-acetaminophen (is it the same drug)
  - Hydrocodone-acetaminophen


Lets get a tally of the unique number of observation for the drugs


```{r}
df_count <- df_split2 %>% count(current_med)

# Remove some special symbol
df_count <- df_count %>%
  mutate(current_med = gsub("\\ ", "", current_med)) %>%
  filter(!current_med == "")

write.csv(df_count, file.path(wd$output, "Observation_counts.csv"))
```


We used chat GPT-3 to find the category of these drugs. The linked search query is [here](https://chat.openai.com/share/392e3a0d-bc7d-44b6-be85-56fb54f4f539). In total we get 68 categories. Read the result and merge with the count


```{r}
df_gpt <- read_xlsx(file.path(wd$data, "Drug_categorized.xlsx"))

# Merge with the count

df_count <- df_count %>% left_join(df_gpt, by = c("current_med" = "Original_Input"))

write.csv(df_count, file.path(wd$output, "Observation_counts.csv"))
```


# Objective 2

**Heatmap of pre-categorized drug**

There is a column in the file 'curated.meds'. This is a manual curation by Araballa's team. We simply make a patient x medicine matrix to draw a heatmap. 

```{r}
list_split <- str_split(dat$curated.meds, ",|;")
names(list_split) <- dat$record_id


un_split <- unlist(list_split, use.names=TRUE)

df_split_main <- stack(list_split) %>%
  # Rename
  dplyr::rename(curated_med = values,
                record_id = ind) 

# Remove NA
df_split_main <- df_split_main %>%
  filter(!curated_med == "NA") 
```



Convert to wide format

```{r}
df_split_main$val <- 1
df_wide <- df_split_main %>% pivot_wider(names_from = curated_med, values_from = val, values_fill = 0) %>%
  mutate(record_id = paste0("p_", record_id))

mat <- as.data.frame(df_wide)
row.names(mat) <- mat$record_id
mat <- mat[,-1]
mat <- as.matrix(mat)
t_mat <- t(mat)
```


Draw a simple Heatmap.


```{r}
mypalette <- c("white", "black")




ph <- 
  Heatmap(t_mat, col=mypalette,
          column_title = "Clustering of patients (columns) based on if a class of drug is given. Each row is a patient",
          show_column_names=FALSE,
          heatmap_legend_param = list(
            color_bar = "discrete"))

pdf(file.path(wd$output, "Heatmap_Curated_cat.pdf"), width = 10, height = 4)
ph
dev.off()
```




























