---
title: "Text cleaning"
date: "2023-09-06"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

# Background

Sammy provided us with a couple of excel and text files documenting the types of drugs administered per-patient. We want to summarize the data. 

# Objectives

1. Summarize all the listed drugs
2. Heatmap of pre-categorized drug


# Pre-processing 

## Loading packages

```{r, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(readxl)
library(ggplot2)
library(reshape2)
library(stringr)
library(ggvenn)

library(ComplexHeatmap)
```


## Directories

```{r}
wd <- list()
wd$main <- here()
wd$data <- file.path(wd$main, "data")
wd$dataP <- file.path(wd$data, "processed")
wd$output <- file.path(wd$main, "output")
wd$script <- file.path(wd$main, "scripts")
```


Create directories

```{r}
if (!file.exists(wd$output )) {
  dir.create(wd$output )
} else {
  print("Directory already exists")
}
```

Load functionm

```{r}
source(file.path(wd$script, "functions.R"))
```


# Objective 1

**Summarize all the listed drugs**

Read in the data

```{r}
dat <- read.delim(file.path(wd$data, "from_Sammy/meds-baseline.tsv"))
head(dat)
```

Each row represents a patient, with the 'record_id' column represeting the patient ID. We extract the 'current_meds_other' column. This is the column detailing what other medication was given. Since multiple drug is given in a column, we separate this.

Sometimes the drug is seperated by a ',' comma or a semi-colon ';'


```{r}
list_split <- str_split(dat$current_meds_other, ",|;")
names(list_split) <- dat$record_id


un_split <- unlist(list_split, use.names=TRUE)

df_split <- stack(list_split) %>%
  # Rename
  dplyr::rename(current_med = values,
                record_id = ind) 

# Remove any empty spaces at the start
df_split <- df_split %>%
  mutate(current_med = gsub("^ ", "", current_med),
         current_med = as.character(mapply(simpleCap, current_med))) %>%
  # Remove blanks
  filter(!current_med == "") %>%
  # Further modification
  mutate(
    # If the start is "-", remove the symbol
    current_med = gsub("^-", "", current_med),
    # Remove leading and trailing whitespace (Does not seem to work as we still have white space)
    current_med = trimws(current_med, which="both", whitespace = "[ \t\r\n]"),
    current_med = trim(current_med)
    )

# # Remove some special symbol
# df_split2 <- df_split %>%
#   mutate(current_med = gsub(" ", " ", current_med))
# #write.csv(df_split2, "~/Desktop/tmp.csv")

```



Quick look at the harmonized category, we see some variation to the names, something like this. 

```{r}
filter(df_split, str_detect(current_med, "Hydrochlorothiazide")) %>% select(current_med) %>% unique()
filter(df_split, str_detect(current_med, "hydrochlorothiazide")) %>% select(current_med) %>% unique()
```

We noted some drugs were separated with a "-" or "/". These indicate muliple drugs was administered. However, not all cases are like this, some "-" is just the name of the drug. So lets have a look at all columns with "-".

```{r}
# Data with "-"
df_split_p1 <- df_split %>% filter(str_detect(current_med, "-"))
tmp <- df_split_p1 %>% pull(current_med) %>% unique()
#write.csv(tmp, "~/Desktop/Data_with_dash.csv")

# Data with "/"
df_split_p2 <- df_split %>% filter(str_detect(current_med, "\\/"))
tmp2 <- df_split_p2 %>% pull(current_med) %>% unique()
#write.csv(tmp2, "~/Desktop/Data_with_slash.csv")

# Data with " "
df_split_p3 <- df_split %>% filter(str_detect(current_med, " "))
tmp3 <- df_split_p3 %>% pull(current_med) %>% unique()
#write.csv(tmp3, "~/Desktop/Data_with_space.csv")
```

We gave this data to Sammy, who then identified which of the drugs are actually multiple drugs. We read the output from Sammy's assessment. It will contain a column to indicate which of these drugs are multiple. 

```{r}
df_dash_ss <-read.csv(file.path(wd$dataP, "Data_with_dash _ss.csv")) %>% 
  select(current_med, multiple_drugs)
  
df_slash_ss <-read.csv(file.path(wd$dataP, "Data_with_slash_ss.csv")) %>%
  select(current_med, multiple_drugs)

df_space_ss <-read.csv(file.path(wd$dataP, "Data_with_space_ss.csv")) %>%
  select(current_med, multiple_drugs)
```


For those that are multiple drugs, we again do the splitting by dash or slash. Lets add back patient names

```{r}
df_dash_ss <- left_join(df_split_p1, df_dash_ss)
df_slash_ss <- left_join(df_split_p2, df_slash_ss)
df_space_ss <- left_join(df_split_p3, df_space_ss)
# Find mulitple drugs, and split them
d1 <- df_dash_ss %>% filter(!multiple_drugs == "")
d2 <- df_slash_ss %>% filter(!multiple_drugs == "")
d3 <- df_slash_ss %>% filter(!multiple_drugs == "")

d_multip <- rbind(d1, d2, d3)

list_split <- str_split(d_multip$multiple_drugs, "-|\\/| ")
names(list_split) <- d_multip$record_id

df_split2 <- stack(list_split) %>%
  # Rename
  dplyr::rename(current_med = values,
                record_id = ind) 


# Lets merge into one big file
# 1) Remove the muliple drugs 
df_no_mult <- df_split %>% filter(!current_med %in% d_multip$current_med)
# 2) Join back with the multiple drugs
df_split_all <- rbind(df_split2, df_no_mult)
# 3) Add back the original medication column (in the unprocessed file)
df_split_all <- left_join(df_split_all, dat) %>% arrange(record_id)

#write.csv(df_split_all, "~/Desktop/current_processed_not_harmonized.csv")
#write.csv(dat, "~/Desktop/original_file_from_ss.csv")
```



## Harmonization

Lets do some harmonization 

```{r}
df_split_all2 <- df_split_all %>%
  mutate(
    # Remove Drug concentration, remove mg
    current_med = gsub("[0-9.-]+mg.*", "", current_med),
    # Remove "Mg" in capital letters with or without a space before a decimal number
    current_med = gsub("([A-Z] )?\\d+(\\.\\d+)? Mg.*", "", current_med),
    # For any drug with open brackets "(", we remove anything after that
    current_med = gsub("\\(.*", "", current_med)
    
    )


```

Lets get a tally of the unique number of observation for the drugs


```{r}
df_count <- df_split_all2 %>% count(current_med)

# Remove some special symbol
df_count <- df_count %>%
  mutate(current_med = gsub("\\ ", "", current_med)) %>%
  filter(!current_med == "")

write.csv(df_count, file.path(wd$output, "Observation_counts_tmpp.csv"))
```

We then input this to chat GPT, 100 lines at a time. Our query is 

> These are list of drugs
>
> Generate 4 column output where ; 
> 1st column is the current input
> 2nd column is the corrected name, for any spelling errors
> 3rd column is the generic / brand name
> 4th column is the drug category


We read the saved output from ChatGP3

```{r}
gpt_output <- read_xlsx(file.path(wd$dataP, "Chat_GPT3_output.xlsx")) %>%
  select(Current_Input:Drug_Category) %>% distinct() 
```

Now we merge the chat GPT3 output with the orignal data

```{r}
df_split_all3 <- df_split_all2 %>%
  mutate(
    current_med = trimws(current_med, which="both", whitespace = "[ \t\r\n]")) %>%
  left_join(gpt_output, by=c("current_med" = "Current_Input"))
```

Check if the number of rows are the same

```{r}
dim(df_split_all2)
dim(df_split_all3)
```


Get the count after harmonization 

```{r}
df_count2 <- df_split_all3 %>% count(harmonize_name)

write.csv(df_count2, file.path(wd$output, "Observation_counts_harmonized.csv"))
```


Get the count for each drug category and generic name

```{r}
drug_cat <-  df_split_all3 %>% count(Drug_Category, Generic_Brand_Name)

write.csv(drug_cat, file.path(wd$output, "Observation_counts_DrugCat_and_Generic_or_BrandName.csv"))


drug_cat <-  df_split_all3 %>% count(harmonize_name, Generic_Brand_Name)

write.csv(drug_cat, file.path(wd$output, "Observation_counts_harmonize_name_and_Generic_or_BrandName.csv"))


drug_cat <-  df_split_all3 %>% count(current_med, harmonize_name, Generic_Brand_Name)

write.csv(drug_cat, file.path(wd$output, "Observation_counts_current_med_and_harmonize_name_and_Generic_or_BrandName.csv"))
```


Save with the patient data as well

```{r}
write.csv(df_split_all3, file.path(wd$output, "Per_patient_harmonized.csv"))
writexl::write_xlsx(df_split_all3, file.path(wd$output, "Per_patient_harmonized.xlsx"))
```


## Sammy input

Sammy helped us harmonizing the data. We read her file. 

```{r}
sam_in <- read_xlsx(file.path(wd$dataP, "round2/harmonized_drug_list_ss.xlsx"))
```
Because excel output something a bit different then R, in terms of the text. We read our saved excel file into R

```{r}
zak_in <- read_xlsx(file.path(wd$output, "Per_patient_harmonized.xlsx"))
```

As we will join according to 'curated_med' column, we check the overlap between Zaki and Sammy files.

```{r}
list_venn <- list(Sammy = sort(sam_in$current_med),
                  Zaki = sort(unique(df_split_all3$current_med)))

ggvenn(list_venn, c("Sammy", "Zaki"))     

#> We manually checked overlap using excel and found 2,303 (of 2,344, 98%) were overlapping. 
```

Based on the Venn, we see 96.5% overlap. So lets work on the overlap first

We filter to only those overlapping. 

```{r}
# only in sammy
only_sam <- sam_in %>% filter(!current_med %in% df_split_all3$current_med)

# only in Zaki
only_zak <- df_split_all3 %>% filter(!current_med %in% sam_in$current_med) %>% arrange(current_med)

# The overlapping 
ov <- df_split_all3 %>% filter(current_med %in% sam_in$current_med) %>% 
  left_join(sam_in)
```

Now we want to harmonize it based on the column provided by Sammy. Because its just 5 columns, lets manually do column-by-column.


```{r}
# Make one column of harmonized name
t1 <- ov %>% mutate(harmonized_final = harmonized_generic_1) %>%
  filter(!is.na(harmonized_final)) %>%
  filter(!harmonized_final == "NA")

t2 <- ov %>% mutate(harmonized_final = harmonized_generic_2) %>%
  filter(!is.na(harmonized_final)) %>%
  filter(!harmonized_final == "NA")

t3 <- ov %>% mutate(harmonized_final = harmonized_generic_3) %>%
  filter(!is.na(harmonized_final)) %>%
  filter(!harmonized_final == "NA")

t4 <- ov %>% mutate(harmonized_final = harmonized_generic_4) %>%
  filter(!is.na(harmonized_final)) %>%
  filter(!harmonized_final == "NA")

t5 <- ov %>% mutate(harmonized_final = harmonized_generic_5) %>%
  filter(!is.na(harmonized_final)) %>%
  filter(!harmonized_final == "NA")


t_all <- rbind(t1, t2, t3, t4, t5) %>% arrange(record_id) %>%
  mutate(harmonized_final = as.character(mapply(simpleCap, harmonized_final)))

writexl::write_xlsx(t_all, file.path(wd$output, "Per_patient_harmonized_round2.xlsx"))

drug_cat <-  t_all %>% count(current_med, harmonized_final) %>% arrange(desc(n))
writexl::write_xlsx(drug_cat, file.path(wd$output, "Tally_current_med_and_harmonized_round2.xlsx"))


drug_cat2 <-  t_all %>% count(harmonized_final) %>% arrange(desc(n))
writexl::write_xlsx(drug_cat2, file.path(wd$output, "Tally_harmonized_round2.xlsx"))

```







# Not done

Lets further harmonize the name

```{r}
df_split2 <- df_split %>%
  mutate(
    current_med = 
      case_when(
        str_detect(current_med, "(?i)hydrochlorothiazide") ~ "Hydrochlorothiazide",
        str_detect(current_med, "(?i)Clonazepam") ~ "Clonazepam",
        str_detect(current_med, "(?i)cymbalta") ~ "Cymbalta",
        str_detect(current_med, "(?i)diphenhydramine") ~ "Diphenhydramine",
        str_detect(current_med, "(?i)gabapentin") ~ "Gabapentin",
        str_detect(current_med, "(?i)hydrocodone") ~ "Hydrocodone",
        str_detect(current_med, "(?i)imodium") ~ "Imodium",
        str_detect(current_med, "(?i)levofloxacin") ~ "Levofloxacin",
        str_detect(current_med, "(?i)ondansetron") ~ "Ondansetron",
        str_detect(current_med, "(?i)prochlorperazine") ~ "Prochlorperazine",
        str_detect(current_med, "(?i)soma") ~ "Soma",
        str_detect(current_med, "(?i)acetaminophen") ~ "Acetaminophen",    # SAME as oxydone?
        str_detect(current_med, "(?i)docusatesodium") ~ "Docusatesodium",
        str_detect(current_med, "(?i)methimazole") ~ "Methimazole",
        str_detect(current_med, "(?i)nicotine") ~ "Nicotine",
        str_detect(current_med, "(?i)oxycodone") ~ "Oxycodone",      # SAME as acetaminophen?
        str_detect(current_med, "(?i)prochlorperazine") ~ "Prochlorperazine",
        str_detect(current_med, "(?i)Coq10") ~ "Coq10",
        str_detect(current_med, "(?i)Aldactazide") ~ "Aldactazide",
        str_detect(current_med, "(?i)Buspirone") ~ "Buspirone",
        str_detect(current_med, "(?i)Cefdinir") ~ "Cefdinir",
        str_detect(current_med, "(?i)Potassium") ~ "Potassium",
        str_detect(current_med, "(?i)levothyroxine") ~ "Levothyroxine",
        str_detect(current_med, "(?i)nystatin") ~ "Nystatin",
        str_detect(current_med, "(?i)clearlax") ~ "Clearlax",
        str_detect(current_med, "(?i)Polyethylene") ~ "Polyethylene Glycol",
        str_detect(current_med, "(?i)Adavir") ~ "Adavir",
        str_detect(current_med, "(?i)Albuterol") ~ "Albuterol",
        str_detect(current_med, "(?i)Hydrochlorothiazide") ~ "Hydrochlorothiazide",
        str_detect(current_med, "(?i)Alfuzosin") ~ "Alfuzosin",
        str_detect(current_med, "(?i)Alfuzusin") ~ "Alfuzosin",
        str_detect(current_med, "(?i)Aleve") ~ "Naproxen",
        str_detect(current_med, "(?i)Alieve") ~ "Naproxen",
        str_detect(current_med, "(?i)Allegra") ~ "Fexofenadine",
        str_detect(current_med, "(?i)Fexofenadine") ~ "Fexofenadine",
        str_detect(current_med, "(?i)Aller-tec") ~ "Cetirizine",
        str_detect(current_med, "(?i)Cetrizine") ~ "Cetirizine",
        str_detect(current_med, "(?i)Allopurinol") ~ "Alloporinol",
        str_detect(current_med, "(?i)Alopurinol") ~ "Alloporinol",
        str_detect(current_med, "(?i)Alogliptin") ~ "Alogliptin",
        str_detect(current_med, "(?i)Aloogliptin") ~ "Alogliptin",
        str_detect(current_med, "(?i)Aloxi") ~ "Palonosetron",
        str_detect(current_med, "(?i)Palonosetron") ~ "Palonosetron",
        str_detect(current_med, "(?i)Alphagan") ~ "Brimonidine",
        str_detect(current_med, "(?i)Brimonidine") ~ "Brimonidine",
        str_detect(current_med, "(?i)Alphalipoiu Acid") ~ "Alpha_Lipoic_Acid",
        str_detect(current_med, "(?i)Alpraz") ~ "Alprazolam",
        str_detect(current_med, "(?i)Altace") ~ "Ramipril",
        str_detect(current_med, "(?i)Ramipril") ~ "Ramipril",
        str_detect(current_med, "(?i)Amaryl") ~ "Glimepiride",
        str_detect(current_med, "(?i)Glimepiride") ~ "Glimepiride",
        str_detect(current_med, "(?i)Ambien") ~ "Zolpidem",
        str_detect(current_med, "(?i)Amelopdipine") ~ "Amlodipine",
        str_detect(current_med, "(?i)Amiodipine") ~ "Amlodipine",
        str_detect(current_med, "(?i)Amilodipine") ~ "Amlodipine",
        str_detect(current_med, "(?i)Amlcolipine") ~ "Amlodipine",
        str_detect(current_med, "(?i)Amlo") ~ "Amlodipine",
        str_detect(current_med, "(?i)Amplodipine") ~ "Amlodipine",
        str_detect(current_med, "(?i)Amidarone") ~ "Amiodarone",
        str_detect(current_med, "(?i)Amidavone") ~ "Amiodarone",
        str_detect(current_med, "(?i)Amoro Ellipta") ~ "Umeclidinium",
        str_detect(current_med, "(?i)Anaro Ellipta") ~ "Umeclidinium",
        str_detect(current_med, "(?i)Anoro") ~ "Umeclidinium",
        str_detect(current_med, "(?i)Anusol") ~ "Hydrocortisone",
        str_detect(current_med, "(?i)Hydrocortisone") ~ "Hydrocortisone",
        str_detect(current_med, "(?i)Apixiban") ~ "Apixaban",
        str_detect(current_med, "(?i)Apriso") ~ "Mesalamine",
        str_detect(current_med, "(?i)Aptiom") ~ "Eslicarbazepine",
        str_detect(current_med, "(?i)Aranesp") ~ "Darbepoetin",
        str_detect(current_med, "(?i)Darbepoetin") ~ "Darbepoetin",
        str_detect(current_med, "(?i)Areds") ~ "Areds",
        str_detect(current_med, "(?i)Arfarin") ~ "Warfarin",
        str_detect(current_med, "(?i)Aricept") ~ "Donepezil",
        str_detect(current_med, "(?i)Arimidex") ~ "Anastrozole",
        str_detect(current_med, "(?i)Arixtra") ~ "Fondaparinux",
        
        str_detect(current_med, "(?i)Aromasin") ~ "Exemestane",
        str_detect(current_med, "(?i)Artorvastatin") ~ "Atorvastatin",
        # ##### str_detect(current_med, "(?i)Asa") ~ "Acetylsalicylic acid,",
        str_detect(current_med, "(?i)Ascorbic Acid") ~ "Ascorbic Acid",
        str_detect(current_med, "(?i)Aspir") ~ "Acetylsalicylic Acid",
        str_detect(current_med, "(?i)Aspririn") ~ "Acetylsalicylic Acid",
        str_detect(current_med, "(?i)Atacand") ~ "Candesartan",
        str_detect(current_med, "(?i)Atenolol") ~ "Atenolol",
        str_detect(current_med, "(?i)Atiran") ~ "Ativan",
        str_detect(current_med, "(?i)Ativan") ~ "Lorazepam",
        str_detect(current_med, "(?i)Ator") ~ "Atorvastatin",
        str_detect(current_med, "(?i)Atonvastatin") ~ "Atorvastatin",
        str_detect(current_med, "(?i)Ativastatin") ~ "Atorvastatin",
        str_detect(current_med, "(?i)Atrova") ~ "Atorvastatin",
        str_detect(current_med, "(?i)Atovastatin") ~ "Atorvastatin",
        str_detect(current_med, "(?i)Atrovent") ~ "Ipratropium",
        # str_detect(current_med, "(?i)") ~ "",
        # str_detect(current_med, "(?i)") ~ "",
        # str_detect(current_med, "(?i)") ~ "",
        # 
        # str_detect(current_med, "(?i)") ~ "",
        
        #str_detect(current_med, "(?i)Albuterol") ~ "Albuterol",
        #str_detect(current_med, "(?i)Allopurinol") ~ "Allopurinol",
        
        TRUE ~ current_med),
    current_med = gsub("[0-9.-]+mg.*", "", current_med),
    # Remove "Mg" in capital letters with or without a space before a decimal number
    current_med = gsub("([A-Z] )?\\d+(\\.\\d+)? Mg.*", "", current_med)
      )

#str_detect(current_med, "(?i)") ~ "",
```


For Aldactazide we want to seperate it to ;
  - Spironolactone 
  - Hydrochlorothiazide

Anaro Ellipta
 - Umeclidinium
 - Vilanterol

Things to check

Drug combination, what to do  
  - Oxycodone-acetaminophen (is it the same drug)  
  - Hydrocodone-acetaminophen 


Lets get a tally of the unique number of observation for the drugs


```{r}
df_count <- df_split2 %>% count(current_med)

# Remove some special symbol
df_count <- df_count %>%
  mutate(current_med = gsub("\\ ", "", current_med)) %>%
  filter(!current_med == "")

write.csv(df_count, file.path(wd$output, "Observation_counts.csv"))
```


We used chat GPT-3 to find the category of these drugs. The linked search query is [here](https://chat.openai.com/share/392e3a0d-bc7d-44b6-be85-56fb54f4f539). In total we get 68 categories. Read the result and merge with the count


```{r}
df_gpt <- read_xlsx(file.path(wd$data, "Drug_categorized.xlsx"))

# Merge with the count

df_count <- df_count %>% left_join(df_gpt, by = c("current_med" = "Original_Input"))

write.csv(df_count, file.path(wd$output, "Observation_counts.csv"))
```


# Objective 2

**Heatmap of pre-categorized drug**

There is a column in the file 'curated.meds'. This is a manual curation by Araballa's team. We simply make a patient x medicine matrix to draw a heatmap. 

```{r}
list_split <- str_split(dat$curated.meds, ",|;")
names(list_split) <- dat$record_id


un_split <- unlist(list_split, use.names=TRUE)

df_split_main <- stack(list_split) %>%
  # Rename
  dplyr::rename(curated_med = values,
                record_id = ind) 

# Remove NA
df_split_main <- df_split_main %>%
  filter(!curated_med == "NA") 
```



Convert to wide format

```{r}
df_split_main$val <- 1
df_wide <- df_split_main %>% pivot_wider(names_from = curated_med, values_from = val, values_fill = 0) %>%
  mutate(record_id = paste0("p_", record_id))

mat <- as.data.frame(df_wide)
row.names(mat) <- mat$record_id
mat <- mat[,-1]
mat <- as.matrix(mat)
t_mat <- t(mat)
```


Draw a simple Heatmap.


```{r}
mypalette <- c("white", "black")




ph <- 
  Heatmap(t_mat, col=mypalette,
          column_title = "Clustering of patients (columns) based on if a class of drug is given. Each row is a patient",
          show_column_names=FALSE,
          heatmap_legend_param = list(
            color_bar = "discrete"))

pdf(file.path(wd$output, "Heatmap_Curated_cat.pdf"), width = 10, height = 4)
ph
dev.off()
```




























